---
title: "Final Project - Lasso Model with Lambda"
author: Hong Fu

format:
    html:
        embed-resources: true
        
---

```{r setup, filename = "Required packages"}
#| message: false
library(tidyverse)
library(caret)
library(glmnet)
library(ISLR)
library(Matrix)
```



# Abstract {.tabset .tabset-fade .tabset-pills}

The following model use Lasso with lambda to 1) dimension reduction statistically 2) cross validate performance results 3) display the performance matrix on accuracy, sensitivity, F1-score, 

# Introduction 

To analyse lasso with lambda approach on shrinking feature space using Azhelmeimer data set using 70/30 ratio using a seed (5003)

```{r, setup1}
Azhelmeimer_data <- read.csv("alzheimers_disease_data.csv")

# Remove rows with NA in Diagnosis 
Azhelmeimer_data <- na.omit(data)

# Using sapply to check the class of each column and retain numeric columns
Azhelmeimer_data <- Filter(is.numeric, data)

# list of features
features <- colnames(Azhelmeimer_data)
features <- features[features != "Diagnosis"]

# Set seed for reproducibility
set.seed(123)

# Train/Test split
train_indices <- createDataPartition(Azhelmeimer_data$Diagnosis, p = 0.6)[[1]]
train_data <- Azhelmeimer_data[train_indices, ]
test_data <- Azhelmeimer_data[-train_indices, ]

# Prepare the data matrices for glmnet
x_train <- as.matrix(train_data[, features])
y_train <- train_data$Diagnosis 
x_test <- as.matrix(test_data[, features])
y_test <- test_data$Diagnosis


```

# Train a Lasso regression model with cross-validation and optimal lambda 

```{r, 3}
# Function to calculate RMSE
calc_rmse <- function(model, x_test, y_test) {
  predictions <- predict(model, x_test)
  sqrt(mean((y_test - predictions)^2))
}

# Initial RMSE with the Lasso model using all features
initial_rmse <- calc_rmse(lasso_model, x_test, y_test)
cat("Initial RMSE with Lasso regression and all features include is :", sprintf("%.2f",initial_rmse), "\n")


```
## Using stepwise to remove 5 less significant features based on RMSE value 



# Method (iii) - lasso approach with function of $\lambda$ 

```{r plot}
# Load necessary libraries
library(glmnet)
library(tidyr)
library(ggplot2)

# Assume x_train and y_train are already defined

# Set a custom lambda grid for a wider range of values
grid <- 10^seq(8, -2, length = 100)

# Perform cross-validation to find the optimal lambda
lasso_cv <- cv.glmnet(x_train, y_train, alpha = 1, lambda = grid)

# Extract the best lambda value (lambda.min) and 1 SE
optimal_lambda <- lasso_cv$lambda.min
optimal_lambda_1se <- lasso_cv$lambda.1se    # Lambda within 1 SE of the minimum

# Fit the Lasso model using the full lambda grid
lasso_model <- glmnet(x_train, y_train, alpha = 1, lambda = grid)
lasso_coefs <- coef(lasso_model)

# Extract the coefficients from the Lasso model as a function of lambda
lasso_coefs <- as.matrix(lasso_coefs)

# Convert the coefficient matrix to a tidy format
coef_df <- as.data.frame(lasso_coefs)

# Create a 'feature' column for each feature name
coef_df$feature <- rownames(coef_df)

# Reshape the data to a long format using pivot_longer
coef_df_long <- coef_df %>%
  pivot_longer(cols = -feature, names_to = "lambda", values_to = "coefficient")

# Add the correct lambda values (repeated for each feature)
coef_df_long$lambda <- rep(grid, each = nrow(coef_df))

# Convert 'lambda' to numeric type (it might be a string after pivot_longer)
coef_df_long$lambda <- as.numeric(coef_df_long$lambda)

# Now create a plot with vertical lines for both optimal_lambda and optimal_lambda_1se
ggplot(coef_df_long, aes(x = log(lambda), y = coefficient, color = feature)) +
  geom_line() +
  labs(title = "Lasso Coefficients as a Function of Lambda",
       x = "Log(Lambda)", y = "Coefficient Value") +
  theme_minimal() +
  
  # Add vertical line for optimal lambda (lambda.min)
  geom_vline(xintercept = log(optimal_lambda), color = "blue", linetype = "dashed") +
  annotate("text", x = log(optimal_lambda), y = max(coef_df_long$coefficient), 
           label = paste("Optimal Lambda (Min):", round(optimal_lambda, 4)),
           hjust = 1.2, color = "blue") +
  
  # Add vertical line for lambda.1se
  geom_vline(xintercept = log(optimal_lambda_1se), color = "red", linetype = "dashed") +
  annotate("text", x = log(optimal_lambda_1se), y = max(coef_df_long$coefficient) * 0.9, 
           label = paste("Lambda (1SE):", round(optimal_lambda_1se, 4)),
           hjust = -0.1, color = "red")

```

The first dotted line (blue) is the optimal lambda (min). The second dotted line (red) is the 1 standard devication from the lambda (min). Both value should be used to detemine the optimum RMSE. In this example, optimal lambda min (blue) is the optimimum Log(Lambda) for best result.

## Model's feature coefficients 

Using the coefficients in the best model above and inspect which coefficients have been shrunk to zero i.e, to be removed to reduce a high dimensonal dataset. This is achieved through `cv.glmnet` will return the model fits for each $\lambda$ value. The coefficients for each $\lambda$ value for each features are calcuated and accessed using `cv.out$glmnet.fit$beta`.

```{r}
# Inspect the coefficients in the best model (at optimal lambda)
best_model_coefficients <- coef(lasso_cv, s = "lambda.min")

# Convert the sparse matrix of coefficients into a data frame for inspection
best_model_coefficients_df <- as.data.frame(as.matrix(best_model_coefficients))

# Rename the columns for better readability
colnames(best_model_coefficients_df) <- c("Coefficient")

# Add the feature names to the data frame
best_model_coefficients_df$Feature <- rownames(best_model_coefficients_df)

# Display the coefficients that have been shrunk to zero
zero_coefficients <- best_model_coefficients_df %>% filter(Coefficient == 0)

# Display the coefficients that have been shrunk to zero
cat("Features with coefficients shrunk to zero:\n")
print(zero_coefficients)

# Display the non-zero coefficients
non_zero_coefficients <- best_model_coefficients_df %>% filter(Coefficient != 0)

cat("\nFeatures with non-zero coefficients:\n")
print(non_zero_coefficients)

```

This model dramatically reduce the model's features from 22 to only 5. 

## Model performance on test data 


Using the optimal trained model we can now predict the salary of the test dataset and calculate the sqaure root of mean squared error.


```{r assessment}
# Use the best CV model to predict the test set Salary
# Use the best CV model (with optimal lambda) to predict the Salary on the test set
test_predictions <- predict(lasso_cv, newx = x_test, s = "lambda.min")

# Calculate the Mean Squared Error (MSE) on the test set
test_mse <- mean((y_test - test_predictions)^2)

# Display the test MSE
cat("Sqaure Root of Mean Squared Error on the test set:", sprintf("%.2f",sqrt(test_mse)), "\n")

```
This model's RMSE is higher than both method (i) and (ii). 


# Conclusion 

Different feature reduction methods, such as stepwise selection, Lasso regression with stepwise, and Lasso regression using lambda as a tuning parameter, can yield significantly different results when determining the optimal model. Statistically, the Lasso approach is more intuitive and easier to explain. Using the third method—Lasso with various lambda values—the coefficients of each feature are calculated through the glmnet() function, with the results represented in a matrix. This matrix can then be used to generate a graph, visually identifying the optimal lambda value by analyzing the resulting data. The coefficients are plotted to determine the best lambda value, which effectively increases data sparsity, as shown in this report. Methods (i) and (ii) produced similar RMSE results **`r sprintf("%.2f",best_rmse)`** in this experiment, whereas method (iii) removed **11** features compared to **5**, though it resulted in a higher RMSE value **`r sprintf("%.2f",sqrt(test_mse))`**. If feature reduction is the priority, method (iii) would be the preferred approach.

