---
title: "Project Plan and EDA"
format:
    html:
        embed-resources: true
editor: visual
---

```{r setup, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
# load or install required packages using pacman
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman"); library(pacman)
pacman::p_load(tidyverse, reshape2, knitr, RColorBrewer, psych, caret, class, randomForest, ranger, corrplot, data.table, DescTools, ggcorrplot, kable, lmtest, glmnet, rmarkdown, styler, tidymodels, vcd, xgboost, lattice, plotly, GGally, cowplot, Hmisc, correlation, heatmaply, quartostamp, fastDummies, mlflow, MLMetrics)
```

```{r variables, echo=FALSE}
alzdata <- readRDS("alzdata.RDS")

# create values by variable type
# target variable
target_var <- "Diagnosis"

# numeric/continuous variables
numeric_vars <- names(alzdata)[sapply(alzdata, is.numeric)]

# binary variables (factor with 2 levels)
binary_vars <- names(alzdata)[sapply(alzdata, function(x) is.factor(x) && length(levels(x)) == 2)]
binary_vars <- setdiff(binary_vars, target_var)

# multi categorical variables (> 2 levels)
multicat_vars <- names(alzdata)[sapply(alzdata, function(x) is.factor(x) && length(levels(x)) > 2)]

# all categorical vars (excluding target_var)
all_catvars <- c(binary_vars, multicat_vars)

# all predictor variables (excluding target_var)
predictor_vars <- setdiff(names(alzdata), target_var)
```

```{r subset, echo=FALSE}
# function to subset data on variable groupings
subset_data <- function(data, vars) {
  return(data[, vars])
}

numeric_data <- subset_data(alzdata, numeric_vars)
binary_data <- subset_data(alzdata, binary_vars)
multicat_data <- subset_data(alzdata, multicat_vars)
```

#### Figure 1. Percentage of frequencies in each class

```{r barchart_classimbalance, echo=FALSE}
library(lattice)
diagnosis_counts <- table(alzdata$Diagnosis)
diagnosis_proportions <- prop.table(diagnosis_counts)
imbalance_barplot <- barplot(diagnosis_proportions * 100,
  col = RColorBrewer::brewer.pal(3, "Set2"),
  ylab = "Percentage",
  ylim = c(0, 100),
  border = NA,
  cex.names = 0.8,
  cex.axis = 0.6,
  cex.lab = 0.6,
  width = 0.1,
  space = 0.1,
  names.arg = c("No Alzheimers", "Alzheimer's")
)

par(mai = c(1, 0.5, 0.5, 0.2))
abline(h = 0, col = "black", lwd = 1)
title(
  main = "Figure 1. Percentage of frequencies in each class (Diagnosis)",
  adj = 0,
  cex.main = 0.8,
  font.main = 1
)
```

In our target variable, the negative class of "No Alzheimer's" has `r table(alzdata$Diagnosis)["No Alzheimer's"]` people, and our positive class of "Alzheimer's" has `r table(alzdata$Diagnosis)["Alzheimer's"]` people. This is a `r round(prop.table(table(alzdata$Diagnosis))["No Alzheimer's"] * 100, 2)`% representation of cases with 'No Alzheimers' (negative class), and a `r round(prop.table(table(alzdata$Diagnosis))["Alzheimer's"] * 100, 2)`% representation of cases with 'Alzheimer's' (positive class) (shown in Figure 1). This suggests that we have **moderate class imbalance** that should be addressed to prevent model bias towards the majority class.

#### Figure 2. Numeric predictor variables by diagnosis

```{r violinplot, echo=FALSE, error=FALSE, warning=FALSE}
library(ggplot2)

# long format
numeric_vars_long <- reshape2::melt(alzdata, id.vars = "Diagnosis", measure.vars = numeric_vars)

# violin plots for numeric variables by target_var
ggviolin <- ggplot(numeric_vars_long, aes(x = Diagnosis, y = value, fill = Diagnosis)) +
  geom_violin(trim = FALSE) +
  facet_wrap(~variable, scales = "free") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Figure 2. Violin plots of numeric predictor variables by diagnosis") +
  theme(
    plot.title = element_text(hjust = 0, size = 8),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
  )

print(ggviolin)

# create plotly from ggplot object
violin_plotly <- ggplotlyr(ggviolin)
```

Figure 2 illustrates the distribution and spread of all the numeric predictor variables grouped by Alzheimer's classification. BMI, Alcohol Consumption, Physical Activity, Diet Quality and all the cholesterol levels appear to be similarly distributed between both groups. MMSE shows more lower scores in the Alzheimer's group, and more higher scores in the no Alzheimer's group. The same pattern is seen in the Functional Assessment and the ADL scores. This suggests they could be important features that impact our classification.

#### Figure 3. Percentage of categorical predictor variable responses by diagnosis group

Figure 3 shows the percentage of categorical predictor variables for both the No Alzheimer's and Alzheimer's classes.

```{r stackedbar, echo=FALSE, warning=FALSE, error=FALSE}
library(GGally)
library(ggplot2)
library(cowplot)

eachvar_plotlist <- list()
for (var in all_catvars) {
  p <- ggally_colbar(alzdata, aes_string(x = target_var, y = var)) +
    theme(
      panel.grid.major.y = element_blank(),
      axis.title.y = element_blank(),
      panel.grid.minor.y = element_blank(),
      legend.position = "none",
      axis.line.y = element_blank(),
    )
  eachvar_plotlist[[var]] <- p
}

for (var in all_catvars) {
  print(eachvar_plotlist[[var]])
}

grid_colbar <- plot_grid(plotlist = eachvar_plotlist, ncol = 4)
print(grid_colbar)
```

#### Figure 4. Correlation heatmap of numeric predictor variables

```{r correlations, echo = FALSE, warning=FALSE}
library(psych)
library(corrplot)

cor_results <- corr.test(alzdata[, numeric_vars], method = "pearson")

#cor matrix
# cor matrix
numeric_cor_matrix <- cor_results$r

# p value matrix
numeric_p_matrix <- cor_results$p

# print(dim(numeric_cor_matrix))
# print(dim(numeric_p_matrix))

# heatmap
corrheatmap <- corrplot(numeric_cor_matrix,
  method = "color",
  type = "lower",
  tl.cex = 0.5,
  tl.col = "black",
  cl.cex = 0.5,
  mar = c(0, 1, 2, 3),
  p.mat = numeric_p_matrix,
  sig.level = 0.05,
  insig = "label_sig",
  pch = 8,
  pch.cex = 1
)

title("Figure 4. Pearson correlation heatmap for numeric predictor variables",
  cex.main = 0.6,
  adj = 0,
  font.main = 1
)
```

Figure 4 shows very weak correlations between all of the numeric predictor variables. The only statistically significant correlations were between Age and Sleep Quality (R = `r round(cor_results$r['Age', 'SleepQuality'], 2)`, p = `r round(cor_results$p['Age', 'SleepQuality'], 4)`), and ADL and Functional Assessment (R = `r round(cor_results$r['ADL', 'FunctionalAssessment'], 2)`, p = `r round(cor_results$p['ADL', 'FunctionalAssessment'], 4)`). However, the weakness of these results indicates there is no multicollinearity between our numeric predictor variables.

## **Performance metrics**

To measure our models performance, we need to take into account the medical context of our classification problem as well as our class imbalance. For output from our logistic regression model, the predicted probability values can be converted into class predictions using a threshold of 0.5. Adding these values with the real values from the test data, we can create a confusion matrix to examine a number of metrics and compare them between models.

The model with the highest *Sensitivity* score will be of particular importance to us as it measures the model that is best at classifying the positive cases correctly. In our context of Alzheimers disease classification, this is extremely important due to the serious implications that could occur in a real-world scenario if our model misclassified someone with Alzheimer's as not having it.

However, other metrics should be considered to get a comprehensive understanding of the model's performance, especially in our data set which has imbalanced classes. The *F-1 score* is another metric we will focus on because it is a good balance between our most important metric, *Sensitivity,* and *precision* (positive predicticted value), and is less sensitive to imbalanced classes than metrics like *Accuracy*.

In addition, we will look at feature importance and compare the results given between random forest models, decision tree models and the logistic regression. This will help us get a better understanding of what features are most influential in the model's predictions, which could also have a clinical relevance.

### KNN model

```{r split}
set.seed(5003)
train_index <- createDataPartition(knnalzdata$Diagnosis, p = 0.7, list = FALSE)
train_data <- knnalzdata[train_index, ]
test_data <- knnalzdata[-train_index, ]

```

```{r tuning}
library(caret)
repeatcv <- trainControl(
  method = "repeatedcv",
  number = 10,
  repeats = 10,
  search = "random", #can also do GRID
  classProbs = TRUE,
  summaryFunction = prSummary, #can also do twoclassSummary
  savePredictions = "all",
)

knn_grid <- expand.grid(k = c(7, 25, 37, 49, 61, 75, 87, 93, 105, 121, 133, 145, 171, 203, 255, 287))
```

```{r train_knn}

set.seed(5003)
knn_model <- train(
  Diagnosis ~ .,
  data = train_data,
  method = "knn",
  tuneGrid = knn_grid,
  trControl = repeatcv,
  preProcess = c("center", "scale")
)


```

```{r predict}
knn_predictions <- predict(knn_model, newdata = test_data)
knn_CM <- confusionMatrix(knn_predictions, test_data$Diagnosis, positive = "Alzheimers")
print(knn_CM)
```
